---
title: "Experimental Design"
author: "Dr. Gordon Wright & Professor Matt Crump"
format: 
  # pdf: default
  html: default
---

# Basics of Experimental Design: IVs, DVs, and Causality on the Golf Links

## Introduction

::: callout-important
-   Understanding IVs, DVs and Design is crucial for research design
-   Golf putting used as a running example for T-Tests, OneWay ANOVA,
    and 2x2 ANOVA
-   Reminder to revisit foundational concepts if struggling
:::

Understanding the basics of experimental design is crucial for
conducting scientific research, especially your Mini-Dissertation this
year. In this review, we'll explore the concepts of Independent
Variables (IVs), Dependent Variables (DVs), and how experiments are
designed to determine causal links. We'll use golf putting as our
running example.

Clarifying Independent and Dependent Variables for Your
Mini-Dissertation:

I want to remind you how crucial it is to fully grasp the concept of
Independent Variables (IVs) and Dependent Variables (DVs). These were
covered in depth last year and are central to the research design of
your mini-dissertation. If you're experiencing challenges now, it may
stem from not being completely confident or competent in distinguishing
between these key elements.

The Independent Variable is what you, as the researcher, manipulate or
categorize, and the Dependent Variable is what you measure as a result
of these manipulations. Whether you're working on an experimental or
correlational study, mastering this distinction is critical.

Please take time to revisit this foundational concept, as a clear
understanding will simplify the complexity of your research and improve
your analysis moving forward.

## Independent Variables (IVs)

### Definition

An Independent Variable is the factor that researchers manipulate or
change in an experiment. It's the presumed cause in a cause-effect
relationship. I remember this by saying "the IV is the variable I
manipulate!" - silly, but effective!

### Characteristics

-   Controlled by the researcher
-   Has two or more levels or conditions
-   Manipulated to observe its effect on the Dependent Variable

### Golf Putting Example

In a golf putting experiment, potential Independent Variables could
include: - Hole size (e.g., small vs. big) - Putting distance (e.g., 5
feet vs. 10 feet) - Crowd noise (e.g., booing vs. cheering) - Incentive
(e.g., win a egg, or lose an egg)

## Dependent Variables (DVs)

### Definition

A Dependent Variable is the outcome that is measured in an experiment.
It's the presumed effect in a cause-effect relationship. I remember this
as "the DV is the **d**ata I collect that **d**epends on the **result**
of the manipulation"

### Characteristics

-   Observed and measured by the researcher
-   Changes in response to the Independent Variable (if a causal
    relationship exists)
-   Quantifiable and measurable in the right way!

### Golf Putting Example

In our golf putting experiment, potential Dependent Variables could
include:

-   Percentage of successful putts

-   Distance from the hole for missed putts

-   Consistency of putting stroke (measured by the purity of the sound
    when hit)

## Experiments to Determine Causal Links (golf joke!)

### Purpose

Experiments are designed to establish a causal relationship between the
IV and DV by controlling for other factors that might influence the
outcome.

### Key Components

1.  Control: Minimizing the influence of extraneous variables
2.  Randomization: Assigning participants or conditions randomly
3.  Replication: Repeating the experiment to ensure reliability

### Steps in Experimental Design

1.  Formulate a hypothesis
2.  Identify and define the IV and DV
3.  Design the experimental procedure
4.  Collect and analyze data
5.  Draw conclusions

# Golf Putting Example (take 1) Between-groups design

## Experiment Overview

-   Independent Variable (IV):
    -   Presence of crowd noise (2 levels: Booing vs. Cheering)
-   Dependent Variable (DV):
    -   Putting accuracy (percentage of successful putts)

## Between-Groups Design Concept

In this between-groups design:

-   We have two separate groups of participants

-   Each group experiences only one condition of the IV

-   We compare the mean scores of the two groups on the DV

{{< pagebreak >}}

## Golf Putting Experiment Example

### Hypotheses

-   Null Hypothesis (H0):
    -   There is no significant difference in putting accuracy between
        the booing and cheering conditions.
-   Alternative Hypothesis (H1):
    -   There is a significant difference in putting accuracy between
        the booing and cheering conditions.

### Method

#### Participants

-   40 amateur golfers (20 per group)
-   Randomly assigned to either the Booing condition or the Cheering
    condition

#### Materials

-   Standard putters and golf balls
-   Putting green with a straight 10-foot putt
-   Audio system for the cheering condition

#### Procedure

1.  Participants are briefed on the task but not told about the specific
    hypothesis
2.  Each participant attempts 20 putts
3.  Booing group: putts while the horrible crowd boos and jeers
4.  Cheering group: putts while the crowd goes wild with excitement

### Data Structure

Our data might look like this:

| Participant | Condition | Success Rate (%) |
|-------------|-----------|------------------|
| 1           | Booing    | 70               |
| 2           | Booing    | 65               |
| ...         | ...       | ...              |
| 20          | Booing    | 75               |
| 21          | Cheering  | 80               |
| 22          | Cheering  | 85               |
| ...         | ...       | ...              |
| 40          | Cheering  | 78               |

### Analysis

We would use an independent samples t-test to analyze this data,
comparing the means of the two groups.

## Conducting an Independent Samples t-test

### Assumptions

1.  The DV is continuous
2.  The IV consists of two categorical, independent groups
3.  There is independence of observations
4.  No significant outliers in either group
5.  The DV is approximately normally distributed for each group
6.  There is homogeneity of variances

### Calculation

We won't go into this here but it might look like this.

```{r}
#| echo: false
#| warning: false


# Load required libraries
library(ggplot2)
library(dplyr)
library(ggpubr)

# Set seed for reproducibility
set.seed(123)

# Generate sample data
booing_data <- data.frame(
  Condition = "Booing",
  SuccessRate = rnorm(20, mean = 70, sd = 5)
)

cheering_data <- data.frame(
  Condition = "Cheering",
  SuccessRate = rnorm(20, mean = 80, sd = 5)
)

# Combine the data
all_data <- rbind(booing_data, cheering_data)

# Perform t-test
t_test_result <- t.test(SuccessRate ~ Condition, data = all_data)

# Calculate means for arrow placement and vertical lines
booing_mean <- mean(booing_data$SuccessRate)
cheering_mean <- mean(cheering_data$SuccessRate)

# Create the plot
p <- ggplot(all_data, aes(x = SuccessRate, fill = Condition)) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("Booing" = "red", "Cheering" = "blue")) +
  labs(title = "Distribution of Golf Putting Success Rates",
       x = "Success Rate (%)",
       y = "Density") +
  theme_minimal() +
  theme(legend.position = "bottom",
        text = element_text(size = 12))  # Increase base text size

# Add vertical lines for means
p <- p + geom_vline(xintercept = booing_mean, color = "black", linetype = "dashed", size = 1) +
         geom_vline(xintercept = cheering_mean, color = "black", linetype = "dashed", size = 1)

# Add mean labels with percentages
p <- p + annotate("text", x = booing_mean, y = 0.08, label = sprintf("Mean: %.2f%%  ", booing_mean), color = "black", hjust = 1, size = 4) +
         annotate("text", x = cheering_mean, y = 0.08, label = sprintf("  Mean: %.2f%%", cheering_mean), color = "black", hjust = 0, size = 4)


# Add double-ended mean difference arrow
p <- p + geom_segment(aes(x = booing_mean, xend = cheering_mean, y = 0.06, yend = 0.06),
                      arrow = arrow(length = unit(0.3, "cm"), ends = "both", type = "closed"), 
                      color = "black", size = 1)

# Add text annotations for mean difference, t-statistic and p-value
p <- p + annotate("text", x = mean(c(booing_mean, cheering_mean)), y = 0.07,
                  label = sprintf("Mean Difference: %.2f", cheering_mean - booing_mean),
                  color = "black", size = 4)
p <- p + annotate("text", x = mean(c(booing_mean, cheering_mean)), y = 0.04,
                  label = sprintf("t(%.0f) = %.2f, p = %.9f", 
                                  t_test_result$parameter, 
                                  t_test_result$statistic,
                                  t_test_result$p.value),
                  color = "black", size = 4)

# Print the plot
print(p)
```

### Interpretation & Reporting

-   If p \< α (typically 0.05), reject the null hypothesis

-   Report the t-statistic, degrees of freedom, p-value, and effect size
    (e.g., Cohen's d)

> There was a significant difference in putting success rates between
> the cheering (M = 79.74%, SD = 4.15) and booing (M = 70.71%, SD =
> 4.86) conditions; t(37) = -6.32, p \< 0.001, d = 2.00. The cheering
> condition resulted in a 9.04 percentage point increase in putting
> success rate compared to the booing condition.

## Advantages and Limitations

### Advantages

-   Avoids order effects and practice effects
-   Each participant only needs to be tested once
-   Suitable for manipulations that cannot be easily reversed, like a
    nice or awful crowd

### Limitations

-   Requires more participants for the same statistical power as a
    within-subjects design
-   Individual differences between groups could affect the results
    -   For example, one group might coincidentally have more skilled
        golfers
-   Each participant provides data for only one condition, potentially
    limiting the richness of the data
-   More susceptible to the influence of outliers due to smaller
    per-condition sample sizes
-   Cannot account for how individuals might respond differently to
    various conditions
-   May be less sensitive to detecting small effects due to
    between-subject variability
-   Randomization may not always perfectly balance out all potential
    confounding variables
-   If the manipulation is obvious, there might be demand
    characteristics where participants guess the hypothesis and alter
    their behavior
-   In our specific golf example, environmental factors (like slight
    changes in wind or green conditions) might affect one group more
    than the other if all tests for one condition are done together

{{< pagebreak >}}

# Golf Putting Example (take 2) Within-subjects design

## Experiment Overview

-   Independent Variable (IV):
    -   Incentive type (2 levels: Win an egg vs. Lose an egg)
-   Dependent Variable (DV):
    -   Putting accuracy (percentage of successful putts)

## Within-Subjects Design Concept

In this within-subjects design:

-   Each participant experiences both conditions of the IV

-   The order of conditions is counterbalanced across participants

-   We compare each participant's performance across the two conditions

## Golf Putting Experiment Example

### Hypotheses

-   Null Hypothesis (H0):
    -   There is no significant difference in putting accuracy between
        the "win an egg" and "lose an egg" conditions.
-   Alternative Hypothesis (H1):
    -   There is a significant difference in putting accuracy between
        the "win an egg" and "lose an egg" conditions.

### Method

#### Participants

-   20 amateur golfers
-   Each participant completes both conditions in a counterbalanced
    order

#### Materials

-   Standard putters and golf balls
-   Putting green with a straight 10-foot putt
-   Eggs as incentives

#### Procedure

1.  Participants are briefed on the task but not told about the specific
    hypothesis
2.  Each participant attempts 20 putts in each condition
3.  "Win an egg" condition: Participant wins an egg for each successful
    putt
4.  "Lose an egg" condition: Participant starts with 20 eggs and loses
    one for each missed putt
5.  Order of conditions is counterbalanced across participants

### Data Structure

Our data might look like this:

| Participant | Win Egg (%) | Lose Egg (%) |
|-------------|-------------|--------------|
| 1           | 75          | 70           |
| 2           | 80          | 75           |
| ...         | ...         | ...          |
| 20          | 85          | 80           |

### Analysis

We would use a paired samples t-test to analyze this data, comparing
each participant's performance across the two conditions.

## Conducting a Paired Samples t-test

### Assumptions

1.  The DV is continuous
2.  The IV consists of two related groups (repeated measures)
3.  No significant outliers in the differences between the two related
    groups
4.  The distribution of the differences should be approximately normally
    distributed

### Calculation

We won't go into this here but it might look like this.

```{r}
#| echo: false
#| warning: false

# Load required libraries
library(ggplot2)
library(dplyr)
library(ggpubr)

# Set seed for reproducibility
set.seed(123)

# Generate sample data
n <- 20
win_egg_data <- rnorm(n, mean = 80, sd = 5)
lose_egg_data <- rnorm(n, mean = 75, sd = 5)

# Create data frame
all_data <- data.frame(
  Participant = 1:n,
  WinEgg = win_egg_data,
  LoseEgg = lose_egg_data
)

# Reshape data for ggplot
all_data_long <- tidyr::pivot_longer(all_data, cols = c(WinEgg, LoseEgg), names_to = "Condition", values_to = "SuccessRate")

# Perform paired t-test
t_test_result <- t.test(all_data$WinEgg, all_data$LoseEgg, paired = TRUE)

# Calculate means
win_egg_mean <- mean(all_data$WinEgg)
lose_egg_mean <- mean(all_data$LoseEgg)

# Create the plot
p <- ggplot(all_data_long, aes(x = SuccessRate, fill = Condition)) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("WinEgg" = "green", "LoseEgg" = "red")) +
  labs(title = "Distribution of Golf Putting Success Rates",
       x = "Success Rate (%)",
       y = "Density") +
  theme_minimal() +
  theme(legend.position = "bottom",
        text = element_text(size = 12))  # Increase base text size

# Add vertical lines for means
p <- p + geom_vline(xintercept = win_egg_mean, color = "black", linetype = "dashed", size = 1) +
         geom_vline(xintercept = lose_egg_mean, color = "black", linetype = "dashed", size = 1)

# Add mean labels with percentages
p <- p + annotate("text", x = win_egg_mean, y = 0.08, label = sprintf("Mean: %.2f%%  ", win_egg_mean), color = "black", hjust = 0, size = 4) +
         annotate("text", x = lose_egg_mean, y = 0.08, label = sprintf("  Mean: %.2f%%", lose_egg_mean), color = "black", hjust = 1, size = 4)

# Add double-ended mean difference arrow
p <- p + geom_segment(aes(x = lose_egg_mean, xend = win_egg_mean, y = 0.06, yend = 0.06),
                      arrow = arrow(length = unit(0.3, "cm"), ends = "both", type = "closed"), 
                      color = "black", size = 1)

# Add text annotations for mean difference, t-statistic and p-value
p <- p + annotate("text", x = mean(c(lose_egg_mean, win_egg_mean)), y = 0.07,
                  label = sprintf("Mean Difference: %.2f", win_egg_mean - lose_egg_mean),
                  color = "black", size = 4)
p <- p + annotate("text", x = mean(c(lose_egg_mean, win_egg_mean)), y = 0.04,
                  label = sprintf("t(%.0f) = %.2f, p = %.9f", 
                                  t_test_result$parameter, 
                                  t_test_result$statistic,
                                  t_test_result$p.value),
                  color = "black", size = 4)

# Print the plot
print(p)
```

### Interpretation & Reporting

-   If p \< α (typically 0.05), reject the null hypothesis

-   Report the t-statistic, degrees of freedom, p-value, and effect size
    (e.g., Cohen's d)

> There was a significant difference in putting success rates between
> the 'win an egg' (M = 79.87%, SD = 5.09) and 'lose an egg' (M =
> 74.76%, SD = 4.70) conditions; t(19) = 3.55, p \< 0.001, d = 0.79. The
> 'win an egg' condition resulted in a 5.96 percentage point increase in
> putting success rate compared to the 'lose an egg' condition.

## Advantages and Limitations

### Advantages

-   Controls for individual differences in golfing ability
-   Requires fewer participants for the same statistical power
-   Each participant provides data for both conditions, allowing for
    richer within-subject comparisons
-   More sensitive to detecting small effects due to reduced error
    variance

### Limitations

-   Potential for order effects (partially addressed through
    counterbalancing)
-   Practice effects may occur as participants become more familiar with
    the task
-   Fatigue effects may influence performance in the second condition
-   Carryover effects from one condition to another (e.g., strategy
    developed in first condition applied to second)
-   Cannot be used for manipulations that have lasting effects (unlike
    the between-subjects design)
-   Participants may guess the hypothesis more easily, potentially
    leading to demand characteristics
-   Analysis can be more complex, especially if order effects need to be
    accounted for
-   In our specific golf example, the psychological impact of winning or
    losing eggs in the first condition might affect performance in the
    second condition

{{< pagebreak >}}

# Bridge: From t-tests to One-Way ANOVA

When we move beyond comparing just two groups, we enter the realm of
Analysis of Variance (ANOVA). The simplest form of ANOVA is the one-way
ANOVA, which is used when we have one independent variable (IV) with
three or more levels.

## One-Way ANOVA: Three Group Comparison

Let's consider an extension of our TeemuBall experiment where we have
three types of crowd reactions:

Independent Variable (IV): Crowd Reaction (3 levels: Booing, Silence,
Cheering) Dependent Variable (DV): Accuracy (percentage of successful
throws)

### Design Concept

-   Three separate groups of participants
-   Each group experiences one type of crowd reaction

### Method

#### Participants

-   60 participants (20 per group)
-   Randomly assigned to one of three conditions

#### Procedure

1.  Participants are briefed on the task
2.  Each participant makes 20 throws in their assigned condition

### Data Structure

| Participant | Crowd Reaction | Accuracy (%) |
|-------------|----------------|--------------|
| 1           | Booing         | 70           |
| 2           | Booing         | 75           |
| ...         | ...            | ...          |
| 21          | Silence        | 80           |
| ...         | ...            | ...          |
| 41          | Cheering       | 85           |
| ...         | ...            | ...          |

### Analysis

-   One-way between-groups ANOVA

### Key Points

1.  **Extension of t-test**: One-way ANOVA is an extension of the
    independent samples t-test, allowing comparison of more than two
    groups.
2.  **Overall F-test**: ANOVA provides an overall F-test to determine if
    there are any significant differences among the groups.
3.  **Post-hoc tests**: If the overall F-test is significant, post-hoc
    tests (e.g., Tukey's HSD) can determine which specific groups differ
    from each other.
4.  **Efficiency**: One-way ANOVA is more efficient than conducting
    multiple t-tests, reducing the risk of Type I errors.

This bridge from t-tests to one-way ANOVA sets the stage for
understanding more complex designs like the 2x2 ANOVA, where we
introduce a second independent variable.

{{< pagebreak >}}

# TeemuBall: 2x2 ANOVA and Three Flavours of Design

## Overview of 2x2 ANOVA

A 2x2 ANOVA (Analysis of Variance) is a powerful statistical technique
that allows us to examine the effects of two independent variables (IVs)
simultaneously on a dependent variable (DV). It extends the concept of a
t-test, which compares means between two groups, to a more complex
design with four groups.

### Components of a 2x2 ANOVA:

1.  **Main Effect of IV1**: The overall effect of the first independent
    variable, averaged across levels of the second IV.
2.  **Main Effect of IV2**: The overall effect of the second independent
    variable, averaged across levels of the first IV.
3.  **Interaction Effect**: The combined effect of both IVs, which may
    be different from what we'd expect based on the main effects alone.

## 2x2 ANOVA Visual Representations

### Basic 2x2 Table Structure

| Incentive | Booing | Cheering |
|-----------|--------|----------|
| Win       | A      | B        |
| Lose      | C      | D        |

### Main Effect of Crowd Reaction

| Incentive | Booing                | Cheering           |
|-----------|-----------------------|--------------------|
| Win       | [**A**]{color="blue"} | [*B*]{color="red"} |
| Lose      | [**C**]{color="blue"} | [*D*]{color="red"} |

-   You'd expect A&C to consistently differ from B&D to show Crowd has a
    general effect

### Main Effect of Incentive

| Incentive | Booing | Cheering |
|-----------|--------|----------|
| Win       | **A**  | **B**    |
| Lose      | *C*    | *D*      |

-   You'd expect A&B to consistently differ from C&D to show Incentive
    has a general effect

### Interaction Effect

| Incentive | Booing | Cheering                |
|-----------|--------|-------------------------|
| Win       | A      | [**B!**]{color="green"} |
| Lose      | C      | D                       |

If Crowd and Incentive both have general effects, where they
**interact** e.g. B (assuming both positive) something magic might
happen there.

If you have trouble grasping this (common) think about Drug
interactions. You shouldn't take Drug 1 and Drug 2 together or your nose
will grow. If you are not taking either drug, fine. If you are taking
only one of the two drugs, fine. But if you are in the condition where
you are taking both drugs... they interact and your nose grows!

These schematic representations help illustrate:

1.  The basic structure of a 2x2 design
2.  How main effects involve comparisons across one variable (ignoring
    the other)
3.  How interaction effects involve the unique combinations of both
    variables

### Advantages over Multiple t-tests:

1.  **Efficiency**: Conducts multiple comparisons in a single analysis,
    reducing the risk of Type I errors.
2.  **Interaction Detection**: Can identify how the effect of one IV
    depends on the levels of the other IV.
3.  **Increased Power**: Generally has more statistical power to detect
    effects than multiple separate t-tests.

## Experiment Overview

Task:

-   Participants throw balls into Teemu's Topmiler (a hat that looks
    like a basketball net)

Independent Variables (IVs):

-   1\. Crowd Reaction: Booing vs. Cheering

-   2\. Incentive: Win an egg vs. Lose an egg

Dependent Variable (DV):

-   Accuracy (percentage of successful throws)

# Fully Between-Groups Factorial Design: 2(b) x 2(b) ANOVA

## Design Concept

-   Four separate (groups of) participants
-   Each group experiences one combination of the two IVs

## Method

### Participants

-   80 participants (20 per group)
-   Randomly assigned to one of four conditions

### Procedure

1.  Participants are briefed on the task
2.  Each participant makes 20 throws in their assigned condition
3.  Conditions:
    -   Booing, Win an egg
    -   Booing, Lose an egg
    -   Cheering, Win an egg
    -   Cheering, Lose an egg

## Data Structure

| Participant | Crowd Reaction | Incentive | Accuracy (%) |
|-------------|----------------|-----------|--------------|
| 1           | Booing         | Win egg   | 75           |
| 2           | Booing         | Win egg   | 80           |
| ...         | ...            | ...       | ...          |
| 21          | Booing         | Lose egg  | 65           |
| ...         | ...            | ...       | ...          |
| 41          | Cheering       | Win egg   | 85           |
| ...         | ...            | ...       | ...          |
| 61          | Cheering       | Lose egg  | 70           |
| ...         | ...            | ...       | ...          |

## Analysis

-   Two-way between-groups ANOVA

## Advantages and Limitations

### Advantages

-   No order effects
-   Each participant only tested once
-   Can examine both main effects and interaction

### Limitations

-   Requires larger sample size
-   Cannot account for individual differences across conditions

# Mixed Design: 2(w) x 2(b) ANOVA

## Design Concept

-   Two (groups of) participants
-   Each participant experiences both levels of Incentive, but only one
    level of Crowd Reaction

## Method

### Participants

-   40 participants (20 per group)
-   Randomly assigned to either Booing or Cheering condition

### Procedure

1.  Participants are briefed on the task
2.  Each participant makes 20 throws in the "Win egg" condition and 20
    throws in the "Lose egg" condition
3.  Incentive order is counterbalanced
4.  Participants remain in their assigned Crowd Reaction condition for
    all throws

## Data Structure

| Participant | Crowd Reaction | Win Egg (%) | Lose Egg (%) |
|-------------|----------------|-------------|--------------|
| 1           | Booing         | 75          | 70           |
| 2           | Booing         | 80          | 75           |
| ...         | ...            | ...         | ...          |
| 21          | Cheering       | 85          | 80           |
| ...         | ...            | ...         | ...          |

## Analysis

-   Mixed ANOVA with Incentive as within-subjects factor and Crowd
    Reaction as between-groups factor

## Advantages and Limitations

### Advantages

-   Combines benefits of within and between-subjects designs
-   Requires fewer participants than fully between-groups
-   Can examine individual differences in response to Incentive

### Limitations

-   Potential for order effects in the Incentive factor
-   More complex analysis
-   Cannot examine individual differences across Crowd Reaction
    conditions

# Mixed Design: 2(b) x 2(w) ANOVA

## Design Concept

-   Two groups of participants
-   Each participant experiences both levels of Crowd Reaction, but only
    one level of Incentive

## Method

### Participants

-   40 participants (20 per group)
-   Randomly assigned to either Win egg or Lose egg condition

### Procedure

1.  Participants are briefed on the task
2.  Each participant makes 20 throws in the Booing condition and 20
    throws in the Cheering condition
3.  Crowd Reaction order is counterbalanced
4.  Participants remain in their assigned Incentive condition for all
    throws

## Data Structure

| Participant | Incentive | Booing (%) | Cheering (%) |
|-------------|-----------|------------|--------------|
| 1           | Win egg   | 75         | 85           |
| 2           | Win egg   | 80         | 90           |
| ...         | ...       | ...        | ...          |
| 21          | Lose egg  | 65         | 75           |
| ...         | ...       | ...        | ...          |

## Analysis

-   Mixed ANOVA with Crowd Reaction as within-subjects factor and
    Incentive as between-groups factor

## Advantages and Limitations

### Advantages

-   Combines benefits of within and between-subjects designs
-   Requires fewer participants than fully between-groups
-   Can examine individual differences in response to Crowd Reaction

### Limitations

-   Potential for order effects in the Crowd Reaction factor
-   More complex analysis
-   Cannot examine individual differences across Incentive conditions

# Repeated Measures Design: 2(w) x 2(w) ANOVA

## Design Concept

-   One group of participants
-   Each participant experiences all combinations of the two IVs

## Method

### Participants

-   20 participants

### Procedure

1.  Participants are briefed on the task
2.  Each participant completes 20 throws in each of the four conditions:
    -   Booing, Win egg
    -   Booing, Lose egg
    -   Cheering, Win egg
    -   Cheering, Lose egg
3.  Order of conditions is counterbalanced

## Data Structure

| Participant | Booing, Win (%) | Booing, Lose (%) | Cheering, Win (%) | Cheering, Lose (%) |
|---------------|---------------|---------------|---------------|---------------|
| 1 | 75 | 70 | 85 | 80 |
| 2 | 80 | 75 | 90 | 85 |
| ... | ... | ... | ... | ... |

## Analysis

-   Two-way repeated measures ANOVA

## Advantages and Limitations

### Advantages

-   Requires fewest participants
-   Can examine individual differences across all conditions
-   Highest statistical power

### Limitations

-   Most susceptible to order effects and practice effects
-   Potential for participant fatigue
-   Most complex analysis
-   Cannot be used if effects of conditions are long-lasting

# Looking at ANOVA main effects and interactions

## Into some detail

::: {.callout-note collapse="true"}
-   Copied with thanks from Matt Crump's Asking Questions with Data OER
    under CC-BY-NC-SA 4.0 Online Textbook:
    https://crumplab.github.io/statistics/

-   Citation: Crump, M. J. C., Navarro, D. J., & Suzuki, J. (2019, June
    5). Answering Questions with Data (Textbook): Introductory
    Statistics for Psychology Students.
    https://doi.org/10.17605/OSF.IO/JZE52
:::

Designs with multiple factors are very common. When you read a research
article you will often see graphs that show the results from designs
with multiple factors. It would be good for you if you were comfortable
interpreting the meaning of those results. The skill here is to be able
to look at a graph and see the pattern of main effects and interactions.
This skill is important, because the patterns in the data can quickly
become very complicated looking, especially when there are more than two
independent variables, with more than two levels.

## 2x2 designs

Let's take the case of 2x2 designs. There will always be the possibility
of two main effects and one interaction. You will always be able to
compare the means for each main effect and interaction. If the
appropriate means are different then there is a main effect or
interaction. Here's the thing, there a bunch of ways all of this can
turn out. Check out the ways, there are 8 of them:

1.  no IV1 main effect, no IV2 main effect, no interaction
2.  IV1 main effect, no IV2 main effect, no interaction
3.  IV1 main effect, no IV2 main effect, interaction
4.  IV1 main effect, IV2 main effect, no interaction
5.  IV1 main effect, IV2 main effect, interaction
6.  no IV1 main effect, IV2 main effect, no interaction
7.  no IV1 main effect, IV2 main effect, interaction
8.  no IV1 main effect, no IV2 main effect, interaction

OK, so if you run a 2x2, any of these 8 general patterns could occur in
your data. That's a lot to keep track of isn't. As you develop your
skills in examining graphs that plot means, you should be able to look
at the graph and visually guesstimate if there is, or is not, a main
effect or interaction. You will need you inferential statistics to tell
you for sure, but it is worth knowing how to know see the patterns.

@fig-10bar22 shows the possible patterns of main effects and
interactions in bar graph form. Here is a legend for the labels in the
panels.

-   1 = there was a main effect for IV1.
-   \~1 = there was **not** a main effect for IV1
-   2 = there was a main effect for IV2
-   \~2 = there was **not** a main effect of IV2
-   1x2 = there was an interaction
-   \~1x2 = there was **not** an interaction

```{r}
#| echo: false
#| label: fig-10bar22
#| fig-cap: "8 Example patterns for means for each of the possible kinds of general outcomes in a 2x2 design."
#| out-width: "100%"



p1 <- data.frame(
  IV1 = c("A", "A", "B", "B"),
  IV2 = c("1", "2", "1", "2"),
  means = c(5, 5, 5, 5)
)

p2 <- data.frame(
  IV1 = c("A", "A", "B", "B"),
  IV2 = c("1", "2", "1", "2"),
  means = c(10, 10, 5, 5)
)

p3 <- data.frame(
  IV1 = c("A", "A", "B", "B"),
  IV2 = c("1", "2", "1", "2"),
  means = c(10, 13, 5, 2)
)

p4 <- data.frame(
  IV1 = c("A", "A", "B", "B"),
  IV2 = c("1", "2", "1", "2"),
  means = c(5, 10, 10, 15)
)

p5 <- data.frame(
  IV1 = c("A", "A", "B", "B"),
  IV2 = c("1", "2", "1", "2"),
  means = c(10, 18, 5, 7)
)

p6 <- data.frame(
  IV1 = c("A", "A", "B", "B"),
  IV2 = c("1", "2", "1", "2"),
  means = c(5, 10, 5, 10)
)

p7 <- data.frame(
  IV1 = c("A", "A", "B", "B"),
  IV2 = c("1", "2", "1", "2"),
  means = c(2, 12, 5, 9)
)

p8 <- data.frame(
  IV1 = c("A", "A", "B", "B"),
  IV2 = c("1", "2", "1", "2"),
  means = c(5, 10, 10, 5)
)

all_22s <- rbind(p1, p2, p3, p4, p5, p6, p7, p8)

type <- c(
  rep("~1, ~2, ~1x2", 4),
  rep("1, ~2, ~1x2", 4),
  rep("1, ~2, 1x2", 4),
  rep("1, 2, ~1x2", 4),
  rep("1, 2, 1x2", 4),
  rep("~1, 2, ~1x2", 4),
  rep("~1, 2, 1x2", 4),
  rep("~1, ~2, 1x2", 4)
)

type <- as.factor(type)

all_22s <- cbind(all_22s, type)

ggplot(all_22s, aes(
  x = IV1,
  y = means,
  group = IV2,
  fill = IV2
)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_classic() +
  facet_wrap( ~ type, nrow = 2) +
  theme(legend.position = "top")
                             
```

@fig-10lines22 shows the same eight patterns in line graph form:

```{r}
#| echo: false
#| label: fig-10lines22
#| fig-cap: "Line graphs showing 8 possible general outcomes for a 2x2 design."
#| out-width: "100%"

ggplot(all_22s, aes(x=IV1, y=means, group=IV2, color=IV2))+
  geom_point()+
  geom_line()+
  theme_classic()+
  facet_wrap(~type, nrow=2)+
  theme(legend.position = "top")
                             
```

The line graphs accentuates the presence of interaction effects.
Whenever the lines cross, or would cross if they kept going, you have a
possibility of an interaction. Whenever the lines are parallel, there
can't be an interaction. When both of the points on the A side are
higher or lower than both of the points on the B side, then you have a
main effect for IV1 (A vs B). Whenever the green line is above or below
the red line, then you have a main effect for IV2 (1 vs. 2). We know
this is complicated. You should see what all the possibilities look like
when we start adding more levels or more IVs. It gets nuts. Because of
this nuttiness, it is often good practice to make your research designs
simple (as few IVs and levels as possible to test your question). That
way it will be easier to interpret your data. Whenever you see that
someone ran a 4x3x7x2 design, your head should spin. It's just too
complicated.

## Interpreting main effects and interactions

The interpretation of main effects and interactions can get tricky.
Consider the concept of a main effect. This is the idea that a
particular IV has a consistent effect. For example, drinking 5 cups of
coffee makes you more awake compared to not drinking 5 cups of coffee.
The main effect of drinking 5 cups of coffee vs not drinking coffee will
generally be true across the levels of other IVs in our life. For
example, let's say you conducted an experiment testing whether the
effect of drinking 5 cups of coffee vs not, changes depending on whether
you are in your house or in a car. Perhaps the situation matters? No,
probably not so much. You will probably still be more awake in your
house, or your car, after having 5 cups of coffee, compared to if you
hadn't.

The coffee example is a reasonably good example of a consistent main
effect. Another silly kind of example might be the main effect of shoes
on your height. For example, if your IV was wearing shoes or not, and
your DV was height, then we could expect to find a main effect of
wearing shoes on your measurement of height. When you wear shoes, you
will become taller compared to when you don't wear shoes. Wearing shoes
adds to your total height. In fact, it's hard to imagine how the effect
of wearing shoes on your total height would ever interact with other
kinds of variables. You will be always be that extra bit taller wearing
shoes. Indeed, if there was another manipulation that could cause an
interaction that would truly be strange. For example, imagine if the
effect of being inside a bodega or outside a bodega interacted with the
effect of wearing shoes on your height. That could mean that shoes make
you taller when you are outside a bodega, but when you step inside, your
shoes make you shorter...but, obviously this is just totally ridiculous.
That's correct, it is often ridiculous to expect that one IV will have
an influence on the effect of another, especially when there is no good
reason.

The summary here is that it is convenient to think of main effects as a
consistent influence of one manipulation. However, when an interaction
is observed, this messes up the consistency of the main effect. That is
the very definition of an interaction. It means that some main effect is
**not** behaving consistently across different situations. Indeed,
whenever we find an interaction, sometimes we can question whether or
not there really is a general consistent effect of some manipulation, or
instead whether that effect only happens in specific situations.

For this reason, you will often see that researchers report their
findings this way:

"We found a main effect of X, BUT, this main effect was qualified by an
interaction between X and Y".

Notice the big **BUT**. Why is it there? The sentence points out that
before they talk about the main effect, they need to first talk about
the interaction, which is making the main effect behave inconsistently.
In other words, the interpretation of the main effect depends on the
interaction, the two things have to be thought of together to make sense
of them.

Here are two examples to help you make sense of these issues:

### A consistent main effect and an interaction

```{r}
#| echo: false
#| label: fig-10mainint
#| fig-cap: "Example means showing a generally consistent main effect along with an interaction"
#| out-width: "100%"
#| fig-asp: 0.5


p7 <- data.frame(
  IV1 = c("A", "A", "B", "B"),
  IV2 = c("1", "2", "1", "2"),
  means = c(2, 12, 5, 9)
)

ggplot(p7, aes(
  x = IV1,
  y = means,
  group = IV2,
  color = IV2
)) +
  geom_point() +
  geom_line() +
  theme_classic() +
  theme(legend.position = "top")

```

@fig-10mainint shows a main effect and interaction. There is a main
effect of IV2: the level 1 means (red points and line) are both lower
than the level 2 means (aqua points and line). There is also an
interaction. The size of the difference between the red and aqua points
in the A condition (left) is bigger than the size of the difference in
the B condition.

**How would we interpret this**? We could say there WAS a main effect of
IV2, BUT it was qualified by an IV1 x IV2 interaction.

**What's the qualification**? The size of the IV2 effect changed as a
function of the levels of IV1. It was big for level A, and small for
level B of IV1.

**What does the qualification mean for the main effect**? Well, first it
means the main effect can be changed by the other IV. That's important
to know. Does it also mean that the main effect is not a real main
effect because there was an interaction? Not really, there is a
generally consistent effect of IV2. The green points are above the red
points in all cases. Whatever IV2 is doing, it seems to work in at least
a couple situations, even if the other IV also causes some change to the
influence.

### An inconsistent main effect and an interaction

```{r}
#| echo: false
#| label: fig-10mainintInc
#| fig-cap: "Example data showing how an interaction exists, and a main effect does not, even though the means for the main effect may show a difference"
#| out-width: "100%"
#| fig-asp: 0.5

p7 <- data.frame(
  IV1 = c("A", "A", "B", "B"),
  IV2 = c("1", "2", "1", "2"),
  means = c(5, 10, 5, 5)
)

ggplot(p7, aes(
  x = IV1,
  y = means,
  group = IV2,
  color = IV2
)) +
  geom_point() +
  geom_line() +
  theme_classic() +
  theme(legend.position = "top")

```

@fig-10mainintInc shows another 2x2 design. You should see an
interaction here straight away. The difference between the aqua and red
points in condition A (left two dots) is huge, and there is 0 difference
between them in condition B. Is there an interaction? Yes!

Are there any main effects here? With data like this, sometimes an ANOVA
will suggest that you do have significant main effects. For example,
what is the mean difference between level 1 and 2 of IV2? That is the
average of the green points ( (10+5)/2 = 15/2= 7.5 ) compared to the
average of the red points (5). There will be a difference of 2.5 for the
main effect (7.5 vs. 5).

Starting to see the issue here? From the perspective of the main effect
(which collapses over everything and ignores the interaction), there is
an overall effect of 2.5. In other words, level 2 adds 2.5 in general
compared to level 1. However, we can see from the graph that IV2 does
not do anything in general. It does not add 2.5s everywhere. It adds 5
in condition A, and nothing in condition B. It only does one thing in
one condition.

What is happening here is that a "main effect" is produced by the
process of averaging over a clear interaction.

**How would we interpret this**? We might have to say there was a main
effect of IV2, BUT we would definitely say it was qualified by an IV1 x
IV2 interaction.

**What's the qualification**? The size of the IV2 effect completely
changes as a function of the levels of IV1. It was big for level A, and
nonexistent for level B of IV1.

**What does the qualification mean for the main effect**? In this case,
we might doubt whether there is a main effect of IV2 at all. It could
turn out that IV2 does not have a general influence over the DV all of
the time, it may only do something in very specific circumstances, in
combination with the presence of other factors.

## Mixed Designs

Throughout this book we keep reminding you that research designs can
take different forms. The manipulations can be between-subjects
(different subjects in each group), or within-subjects (everybody
contributes data in all conditions). If you have more than one
manipulation, you can have a mixed design when one of your IVs is
between-subjects and one of the other ones is within-subjects.
